Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-01 paper.tex
Fixed model for cluster reads

The buffer might be a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up for the two first blocks in a reconstructed image.Implementation details about the
buffer are given later.
\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/buffer.pdf_tex}
\caption{Illustration of the buffer used in cluster reads, showing how
  the two first 4x4 blocks of a reconstructed image would be arranged
  in the buffer.  White portions in the buffer are not allocated.  The
  column-major order is used, similar to the NIfTI format: i is the
  fastest changing dimension, followed by j, followed by k.}
\label{fig:cluster-reads-buffer}
% Data file: path
% Gnuplot file: path
% Script file: path
\end{figure}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written without any seeking. In the worst case, the
memory load only partially covers rows in the reconstructed image:
O($d^2$) seeks are then required during writing, one for every partial
row in every partial slice. In the intermediary case, rows are
complete but some slices can only be partially reconstructed: O($d$)
seeks are then required.

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), complete block rows (case
2) or complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory load to a point that the overall number of seeks also
increases. However, such cases are a bit pathological and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\hfill
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}
To summarize, the amount of memory used $m'$ is set as follows in each of the 3 cases:
%% \begin{equation*}
%%   m_{case 1}' = bd^3\floor*{\frac{m}{d^3b}}; m_{case 2}' = bd^3\sqrt[3]{n}\floor*{\frac{m}{d^3b\sqrt[3]{n}}}; m_{case 3}' = bd^3\sqrt[3]{n}^2\floor*{\frac{m}{d^3b\sqrt[3]{n}^2}}
%% \end{equation*}
%% which gives:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}

In addition, our algorithm avoids configurations where the memory load
overlaps multiple block slices in case 2 or multiple block rows in
case 1, as such overlaps multiply the number of required seeks
(see Figure~\ref{fig:avoided-configurations}-Right). Thus, the total
number of memory loads $x$ required to reconstruct the image is:
\begin{equation*}
  x_1 = \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{n}^2;
  x_2 = \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{n};
  x_3 = \ceil*{\frac{Rb}{m_3'}}
\end{equation*}

Finally, the total number of seeks performed by cluster reads to reconstruct the image is:
%% \begin{equation*}
%% N_\mathrm{cr} =
%% \begin{cases}
%%   n + \sqrt[3]{n}^2 \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}d^2 & \text{if } \frac{m}{d^3b} < \sqrt[3]{n}\\
%%   n + \sqrt[3]{n} \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}d      & \text{if } \sqrt[3]{n} \leq \frac{m}{d^3b} < \sqrt[3]{n}^2\\
%%   n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{m}{d^3b} < n\\
%%   n + 1                                                          & \text{otherwise}
%% \end{cases} \label{eq:seeks-cluster-reads}
%% \end{equation*}
%% And since $d=\frac{D}{\sqrt[3]{n}}$ and $R=D^3$, we finally have:
\begin{equation}
N_\mathrm{cr} =
\begin{cases}
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{R}^2     & \text{if } \frac{mn}{Rb} < \sqrt[3]{n}\\
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{R}        & \text{if } \sqrt[3]{n} \leq \frac{mn}{Rb} < \sqrt[3]{n}^2\\
  n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{mn}{Rb} < n\\
  n + 1                                                          & \text{otherwise}
\end{cases} \label{eq:seeks-cluster-reads-1}
\end{equation}
\todo{Is $N_{cr}$ a continuous function of m?}

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 3)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
true.  Function \texttt{sizeof} (line 8) returns the actual memory
used by its argument, including only its allocated segments in case
the argument is a buffer. It should be noted that writing the buffer
to the reconstructed image (line 11) might involve some manipulation
in the buffer to merge its continuous segments. \todo{Precise this
  based on Valerie's final code.}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE sorted\_splits = sort splits by increasing (z,y,x) values
    \STATE initialize buffer
    \STATE (m',case)=\texttt{switch}(m,n,R,b)
    \STATE old\_block = list[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = list[i]
      \STATE overlap = \texttt{check\_overlap}(block,old\_block,case)
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(block) $\geq$ m' \textbf{or} overlap=true}
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
      \STATE read block and insert it in buffer
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[h]
  \caption{Buffered merging of blocks with multiple reads}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_splits = sort splits by increasing (z,y,x) values
  \STATE start\_index = 0 ; end\_index=(m-1)
  \STATE write\_range = (start\_index, end\_index)
  \WHILE{end\_index \textless Rb}
    \STATE initialize buffer
    \FOR{split in sorted\_splits}
      \IF{split has voxels in write\_range}
        \STATE split\_data = read split
        \STATE in split\_data, extract the rows in write\_range
        \STATE write rows to buffer
      \ENDIF
    \ENDFOR
    \STATE write buffer to reconstructed\_image
    \STATE start\_index = end\_index + 1
    \STATE end\_index += m
  \ENDWHILE

@@

The total number of seeks performed by multiple reads in case $i$ is:
\begin{eqnarray*}
  N^i_{\mathrm{MR}} &=& x_i + (x_i-1)b_i +b_i', \quad i \in \llbracket 1,5 \rrbracket\\
  &=& x_i \left(1+b_i\right)-b_i+b_i'
\end{eqnarray*}
where $x_i$ is the total number of required memory loads, $b_i$ is the
number of blocks accessed by the first ($x_i$-1) memory loads and
$b_i'$ is the number of blocks access by the last memory load. We have:
\begin{equation*}
  x_i = \ceil*{\frac{Rb}{m_i'}}, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}
and:
\begin{equation*}
b_1=k_1 ; \quad b_2=\sqrt[3]{n} ; \quad b_3 = k_3\sqrt[3]{n} ; \quad b_4=\sqrt[3]{n}^2 ; \quad b_5=k_5\sqrt[3]{n}^2
\end{equation*}
and:
\begin{eqnarray*}
b_1'=\sqrt[3]{n}D^2\text{ mod }k_1; &b_2'=b_2;   & b_3' = \sqrt[3]{n}\left( \sqrt[3]{n}D\text{ mod }k_3\right)\\
                                   &b_4'=b_4;  & b_5'=\sqrt[3]{n}^2\left( \sqrt[3]{n}\text{ mod }k_5\right) 
\end{eqnarray*}
It gives the following expression for $N_\mathrm{MR}$:
\begin{equation*}
  N_\mathrm{MR} =
\begin{cases}
  \ceil*{\frac{Rb}{m_1'}}(k_1+1)-k_1
  + \left( \sqrt[3]{n}D^2 \text{ mod } k_1 \right)& \text{\hspace*{-0.9em}if } d \leq \frac{m}{b} < D\\[0.4em]

>>>>

The buffer might be a slightly complex data structure, for instance an
associative array or a Python dictionary, capable of storing multiple
disjoint sequences of contiguous bytes without having to allocate
memory for the bytes between such
sequences. Figure~\ref{fig:cluster-reads-buffer} illustrates how the
buffer would fill up for the two first blocks in a reconstructed
image. Implementation details about the buffer are given later.
\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/buffer.pdf_tex}
\caption{Illustration of the buffer used in cluster reads, showing how
  the two first 4x4 blocks of a reconstructed image would be arranged
  in the buffer.  White portions in the buffer are not allocated.  The
  column-major order is used, similar to the NIfTI format: i is the
  fastest changing dimension, followed by j, followed by k.}
\label{fig:cluster-reads-buffer}
% Data file: path
% Gnuplot file: path
% Script file: path
\end{figure}

The number of seeks performed by cluster reads depends on how blocks
loaded in memory arrange in the reconstructed image. In the best case,
complete contiguous slices of the reconstructed image can be assembled
in memory and written in a single seek. In the worst case, the
memory load only partially covers rows in the reconstructed image:
O($d^2$) seeks are then required during writing, one for every partial
row in every partial slice. In the intermediary case, rows are
complete but some slices can only be partially reconstructed: O($d$)
seeks are then required.

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), of complete block rows (case
2) or of complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory load to a point that the overall number of seeks also
increases. Such cases are, however, slightly unusual and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/incomplete-rows.pdf_tex}
\hfill
\def\svgwidth{0.3\columnwidth}
\input{figures/svg/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 3)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
true.  Function \texttt{sizeof} (line 8) returns the actual memory
used by its argument, including only its allocated segments in case
the argument is a buffer. It should be noted that writing the buffer
to the reconstructed image (line 11) might involve some manipulation
in the buffer to merge its continuous segments. \todo{Precise this
  based on Valerie's final code.}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE sorted\_splits = sort splits by increasing (z,y,x) values
    \STATE initialize buffer
    \STATE (m',case)=\texttt{switch}(m,n,R,b)
    \STATE old\_block = sorted\_splits[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = sorted\_splits[i]
      \STATE overlap = \texttt{check\_overlap}(block,old\_block,case)
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(block) $\geq$ m' \textbf{or} overlap=true}
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
      \STATE read block and insert it in buffer
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

The amount of memory used $m'$ is set as follows in each of the 3 cases:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}
The number of seeks performed by cluster reads in each of the three cases is:
\begin{equation*}
  N^i_{\mathrm{CR}} = n + x_ib_i, \quad i \in \llbracket 1, 3\rrbracket,
\end{equation*}
where $x_i$ is the number of memory loads required to reconstruct the
image and $b_i$ is the number of seeks required to write a the memory
load. The first $n$ seeks in the equation are required to read all the blocks
once. According to Figure~\ref{fig:cluster-reads}, we have:
\begin{equation*}
  b_1=d^2=\sqrt[3]{\frac{R}{n}}^2 \quad ; \quad b_2=d=\sqrt[3]{\frac{R}{n}} \quad ; \quad b_3=1
\end{equation*}

Our algorithm avoids configurations where the memory load
overlaps multiple block slices in case 2 or multiple block rows in
case 1, as such overlaps multiply the number of required seeks
(see Figure~\ref{fig:avoided-configurations}-Right). Thus, the
numbers of memory loads required to reconstruct the image are:
\begin{equation*}
  x_1 = \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{n}^2;
  x_2 = \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{n};
  x_3 = \ceil*{\frac{Rb}{m_3'}}
\end{equation*}

Finally, the total number of seeks performed by cluster reads to reconstruct the image is:
\begin{equation}
N_\mathrm{CR} =
\begin{cases}
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{R}^2     & \text{if } \frac{mn}{Rb} < \sqrt[3]{n}\\[0.4em]
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{R}        & \text{if } \sqrt[3]{n} \leq \frac{mn}{Rb} < \sqrt[3]{n}^2\\[0.4em]
  n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{mn}{Rb} < n\\[0.4em]
  n + 1                                                          & \text{otherwise}
\end{cases} \label{eq:seeks-cluster-reads-1}
\end{equation}
It should be noted that $N_{CR}$ is not a continuous function of m,
due to the differences among $b_i$ values.

@@

\begin{algorithm}[h]
  \caption{Buffered merging of blocks with multiple reads}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_splits = sort splits by increasing (z,y,x) values
  \STATE start\_index = 0 ; end\_index=(m-1)
  \STATE write\_range = (start\_index, end\_index)
  \WHILE{end\_index \textless Rb}
    \STATE initialize buffer
    \FOR{split in sorted\_splits}
      \IF{split has voxels in write\_range}
        \STATE split\_data = read split
        \STATE in split\_data, extract the rows in write\_range
        \STATE insert rows in buffer
      \ENDIF
    \ENDFOR
    \STATE write buffer to reconstructed\_image
    \STATE start\_index = end\_index + 1 ; end\_index += m
  \ENDWHILE

@@

The total number of seeks performed by multiple reads in case $i$ is:
\begin{eqnarray*}
  N^i_{\mathrm{MR}} &=& x_i + (x_i-1)b_i +b_i', \quad i \in \llbracket 1,5 \rrbracket\\
  &=& x_i \left(1+b_i\right)-b_i+b_i'
\end{eqnarray*}
where $x_i$ is the total number of required memory loads, $b_i$ is the
number of blocks accessed by the first ($x_i$-1) memory loads and
$b_i'$ is the number of blocks access by the last memory load. The
first $x_i$ seeks in the equation are required to write each memory
load (1 seek per memory load). We have:
\begin{equation*}
  x_i = \ceil*{\frac{Rb}{m_i'}}, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}
and:
\begin{equation*}
b_1=k_1 ; \quad b_2=\sqrt[3]{n} ; \quad b_3 = k_3\sqrt[3]{n} ; \quad b_4=\sqrt[3]{n}^2 ; \quad b_5=k_5\sqrt[3]{n}^2
\end{equation*}
and:
\begin{eqnarray*}
b_1'=\sqrt[3]{n}D^2\text{ mod }k_1; &b_2'=b_2;   & b_3' = \sqrt[3]{n}\left( \sqrt[3]{n}D\text{ mod }k_3\right)\\
                                   &b_4'=b_4;  & b_5'=\sqrt[3]{n}^2\left( \sqrt[3]{n}\text{ mod }k_5\right) 
\end{eqnarray*}
It gives the following expression for $N_\mathrm{MR}$:
\begin{equation*}
  N_\mathrm{MR} =
\begin{cases}
  \ceil*{\frac{Rb}{m_1'}}(k_1+1)-k_1 \\
  + \left( \sqrt[3]{n}D^2 \text{ mod } k_1 \right)& \text{\hspace*{-0.9em}if } d \leq \frac{m}{b} < D\\[0.4em]
